First steps in synthetic algebraic geometry

### 2017-02-05 slides-warwick2017.tex
Beginning of the slides for the algebraic geometry seminar in Warwick

\title{Using the internal language of toposes in algebraic geometry}
\author{Ingo Blechschmidt}
\date{May 24th, 2016}

@@

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(-0.5cm,0.8cm)},fill=blue!20] at ($(#1.north)+(1.1cm,-1.4cm)$) {#2};}%
}%

@@

  \hil{Using the internal language of toposes in \\ algebraic geometry}
  \medskip

@@

  Category Theory Seminar \\
  Centre for Mathematical Sciences \\
  University of Cambridge \\
  \ \\
  May 24th, 2016
  \par
\end{frame}

\backupstart
{\usebackgroundtemplate{\includegraphics[width=\paperwidth]{images/cambridge-university-botanic-garden}}
\begin{frame}[plain]
  \vspace{\textheight}\vspace{-1em}
  \centering
  \emph{Cambridge University Botanic Garden} \\
  \par
\end{frame}}

\note{
  Photo source:
  \href{http://www.stpaulscambridge.org.uk/classes-and-room-hire/activities-and-events/newtown-news/}{Newtown
  News}
}

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}
\backupend

@@

  \begin{changemargin}{2.5em}{2.5em}
    We describe how the internal language of certain
    toposes, the associated small and big Zariski toposes of a scheme, can be
    used to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.
    % This is useful for studying schemes from a local or relative point of view.

    The starting point is that, from the internal point of view, sheaves of rings
    and sheaves of modules look just like plain rings and plain modules.
    In this way, some concepts and statements of scheme theory can be reduced to
    concepts and statements of intuitionistic linear algebra.

    Furthermore, modal operators can be used to model phrases such as ``on a
    dense open subset it holds that'' or ``on an open neighbourhood of a given
    point it holds that''. These operators define certain subtoposes; a
    generalization of the double-negation translation is useful in order to
    understand the internal universe of those subtoposes from the internal point
    of view of the ambient topos.

    A particularly interesting task is to internalise the
    construction of the relative spectrum, which, given a quasicoherent sheaf of algebras
    on a scheme~$X$, yields a scheme over~$X$. From the internal point of
    view, this construction should simply reduce to an intuitionistically sensible
    variant of the ordinary construction of the spectrum of a ring, but it turns
    out that this expectation is too naive and that a refined approach is
    necessary.

    We also discuss how the little Zariski topos can be described using the
    internal language of the big Zariski topos, and vice versa; here too there
    is a small surprise.
  \end{changemargin}
}

\section{Basics}

\subsection{What is a scheme?}

\begin{frame}\frametitle{What is a scheme?}
  \begin{itemize}
  \item A \hil{manifold} is a space which is \hil{locally isomorphic} to some open
  subset of some~$\RR^n$.
  \medskip
  \item A \hil{scheme} is a space which is \hil{locally isomorphic} to the
  \hil{spectrum of some (commutative) ring}:
  \[ \Spec A \defeq \{ \ppp \subseteq A \,|\, \text{$\ppp$ is a prime ideal} \} \]
  \item By \hil{space} we mean: topological space~$X$ equipped with a local
  sheaf~$\O_X$ of rings.
  \end{itemize}

  \centering
  \begin{columns}
    \begin{column}{0.45\textwidth}
      \centering
      \includegraphics[height=0.25\textheight]{images/manifold} \\
      \tiny a manifold
      \par
    \end{column}
    \begin{column}{0.45\textwidth}
      \centering
      \includegraphics[height=0.25\textheight]{images/mumfords-treasure-map} \\
      \tiny Mumford's treasure map of~$\Spec \ZZ[X]$
      \par
    \end{column}
  \end{columns}
\end{frame}

\note{\justifying
  A \emph{sheaf of rings} on a topological space~$X$ is a ring object
  in~$\Sh(X)$, the category of set-valued sheaves on~$X$.

  A sheaf~$\O_X$ of rings is \emph{local} if and only if all the
  stalks~$\O_{X,x}$ are local rings. Why not demand that the sets of
  sections~$\O_X(U)$ are local rings? This choice has a geometric meaning, but can
  also be motivated from a logical point of view: A sheaf of rings is local if
  and only if, from the point of view of the internal language of~$\Sh(X)$, it
  is a local ring.

  Think of~$\O_X$ as the sheaf of ``number-valued functions'' on~$X$. In
  algebraic geometry, this structure sheaf is a crucial part of the data:
  Wildly different schemes can have the same underlying topological space.
  \par
}

\begin{frame}\frametitle{Motivating the spectrum}
  Let~$A$ be a commutative ring (in~$\Set$).

  Is there a \hil{free local ring}~$A \to A'$ over~$A$?
  \[ \xymatrix{
    A \ar[rd] \ar[rrr] &&& {\substack{\text{local}\\\text{\normalsize$R$}\\\phantom{\text{local}}}} \\
    & {\substack{\text{\normalsize$A'$}\\\text{local}}} \ar@{-->}_[@!34]{\text{local}}[rru]
  } \]

  \hil{No,} if we restrict to~$\Set$.

  \hil{Yes,} if we allow a change of topos:
  Then $A \to \O_{\Spec A}$ is the universal localization.
\end{frame}

\note{\justifying
  Details on this point of view can be found in one of Peter Arndt's very nice
  answers on MathOverflow:
  \begin{center}\url{http://mathoverflow.net/a/14334/31233}\end{center}
  \par
}

\subsection{What is a topos?}

\begin{frame}\frametitle{What is a topos?}
  \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier.
  \end{block}

  \begin{block}{Motto}A topos is a category which is sufficiently rich to support
  an \hil{internal language}.
  \end{block}

  \begin{block}{Examples}\begin{itemize}
    \item \hil{$\Set$:} \tabto{1.4cm} category of sets
    \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
  \end{itemize}\end{block}
\end{frame}

\note{\justifying
  While technically correct, the formal definition is actually
  misleading in a sense: A topos has lots of other vital structure, which
  is crucial for a rounded understanding, but is not listed in the
  definition (which is trimmed for minimality).

  A more comprehensive definition is: A \emph{topos} is a locally cartesian
  closed, finitely complete and cocomplete Heyting category which is exact,
  extensive and has a subobject classifier.

  Check out an
  \href{https://ncatlab.org/publications/published/Leinster2011}{article by
  Tom Leinster} for a leisurely introduction to topos theory.
  \par
}

\subsection{What is the internal language?}

\frame[t]{\frametitle{What is the internal language?}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\E$ \\
      \midrule
      object of~$\E$ & set/type \\
      morphism in~$\E$ & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      group object & group \\
      \bottomrule
    \end{tabular}
  \end{center}
}

@@

\note{
  \begin{itemize}
    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
    formulated over any topos (not just sheaf topoi). They are not all
    arbitrary: Rather, they are very finely concerted to make the crucial
    properties about the internal language (see next slide) true.\end{justify}

@@

  \begin{block}{Crucial property: Soundness}
    If~$U \models \varphi$ and $\varphi$
    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no AxC
    }},
    then~$U \models \psi$.
  \end{block}

  \vfill

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{2.75cm} iff $f|_U = 0 \in \F(U)$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{2.75cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

\note{\scriptsize%
  \vspace{-0.5em}%
  \begin{center}\large\bf Why is constructive mathematics
  interesting?\end{center}
  \vspace{-1em}
  \begin{itemize}
    \item The internal logic of most toposes is constructive.

@@

\note{\scriptsize%
  \vspace{-0.5em}%
  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
  \vspace{-1em}

  \begin{enumerate}
    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
    the term \emph{contradiction} is generally forbidden. This is not the
    case, one has to distinguish between
    \begin{itemize}
      \item\scriptsize a true proof by contradiction: ``Assume~$\varphi$ were false.
      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
    \end{itemize}
    which constructively is only a proof of the weaker
    statement~$\neg\neg\varphi$, and
    \begin{itemize}
      \item\scriptsize a proof of a negated formula: ``Assume~$\psi$ were true. Then
      \ldots, contradiction. So~$\neg\psi$ holds.''
    \end{itemize}
    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}

@@

\note{
  \begin{enumerate}
    \item[3.]\scriptsize \begin{justify}There is one last false rumour about constructive mathematics: Namely
    that most of mathematics breaks down in a constructive setting. This is
    only true if interpreted naively: Often, already very small changes to the
    definitions and statements (which are classically simply equivalent
    reformulations) suffice to make them constructively acceptable.

    In other cases, adding an additional hypothesis, which is classically
    always satisfied, is necessary (and interesting). Here is an example: In
    constructive mathematics, one can not show that any inhabited subset of the
    natural numbers possesses a minimal element. [One can also not show the
    negation -- recall the previous false rumour.] But one can show (quite
    easily, by induction) that any inhabited and \emph{detachable} subset
    of the natural numbers possesses a minimal element. A subset~$U \subseteq
    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
    \in U$ or~$n \not\in U$.

@@

  \begin{justify}\scriptsize References about constructive mathematics include:
  \begin{itemize}
  \item Bridges. Constructive Mathematics.
  \item van Dalen. Intuitionistic logic.
  \item Troelstra and van Dalen. Constructivism in Mathematics: An Introduction.
  \end{itemize}
  \href{http://math.andrej.com/category/constructive-math/}{Andrej Bauer's
  blog} is also very informative.\end{justify}
}

\section{The little Zariski topos of a scheme}
\subsection{Basic look and feel}

@@

\begin{frame}\frametitle{Building a dictionary}
  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set/type \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      % coherent sheaf & coherent module \\
      tensor product of sheaves & tensor product of modules \\
      % sheaf of rational functions & total quotient ring of~$\O_X$ \\
      sheaf of Kähler differentials & module of Kähler differentials \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      dimension of $X$ & Krull dimension of~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

@@

\begin{frame}[c]\frametitle{Praise for Mike Shulman}
  \centering
  \includegraphics[scale=0.4]{images/mike-shulman-stack-semantics}
  \par
\end{frame}

\note{\justifying
  The internal language of a topos supports
  \begin{itemize}
  \item first-order logic,
  \item higher-order logic (for instance quantification over subsets),
  \item dependent types, and
  \item unbounded quantification.
  \end{itemize}

  The first three items are standard. The fourth is due to Mike Shulman.
  Combined, it's possible to interpret ``essentially all of constructive
  mathematics'' internal to a topos.

  Restrictions persist for operations with a ``set-theoretical flavor'' like
  building an infinite union of iterated powersets, for example~$\bigcup_{n \in
  \NN} P^n(\NN)$.
  \par
}

\subsection{The rank function of sheaves of modules}

\begin{frame}\frametitle{The rank function of sheaves of modules}
  There is the following one-to-one correspondence:
  \[\small
    \underunbrace{\fmini{0.35\textwidth}{upper semi-continuous \\ functions on~$X$}}_{\text{external}}
    \xleftrightarrow{\quad\text{1:1}\quad}
    \underunbrace{\fmini{0.4\textwidth}{completed natural numbers}}_{\text{internal}}
  \]

  \vfill

  \begin{center}
    \small
    \begin{minipage}{0.42\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$M$ be a f.\,g.\@ $A$-module. Assume that $A$ is a field.
        Then~$M$ is free iff the minimal number of generators is an
        actual natural number.
      \end{exampleblock}
    \end{minipage}
    \quad\scalebox{3}{$\Rightarrow$}\quad
    \begin{minipage}{0.37\textwidth}
      \begin{exampleblock}{}
        \justifying
        Let~$\F$ be an~$\O_X$-module of finite type. Assume that~$X$ is reduced.
        Then~$\F$ is locally free iff its rank is locally constant.
      \end{exampleblock}
    \end{minipage}
  \end{center}
\end{frame}

\note{
  \begin{block}{Proposition}
    If every inhabited subset of the natural numbers has a minimum, then the
    law of excluded middle holds. (So in constructive mathematics, one cannot
    prove the natural numbers to be complete in this sense.)
  \end{block}

  \begin{block}{Proof}
    Let~$\varphi$ be an arbitrary formula. Define the subset
    \[ U \defeq \{ n \in \NN \,|\, n = 1 \vee \varphi \} \subseteq \NN, \]
    which surely is inhabited by~$1 \in U$. So by assumption, there exists a
    number~$z \in \NN$ which is the minimum of~$U$. We have
    \[ z = 0 \quad\vee\quad z > 0 \]
    (this is constructively not trivial, but can be proven by induction).

    If~$z = 0$, we have~$0 \in U$, so~$0 = 1 \vee \varphi$, so~$\varphi$
    holds.

    If~$z > 0$, then~$\neg\varphi$ holds: Because if~$\varphi$ were
    true, zero would be an element of~$U$, contradicting the minimality of~$z$.
  \end{block}
}

\note{
  \begin{block}{Proposition}
    The partially ordered set
    \[ \widehat\NN \defeq \{ A \subseteq \NN \,|\, \text{$A$ inhabited and upward
    closed}
    \} \]
    is the least partially ordered set containing~$\NN$ and possessing minima
    of arbitrary inhabited subsets.
  \end{block}

  The embedding $\NN \hookrightarrow \widehat\NN$ is given by
  \[ n \in \NN \longmapsto {\uparrow}(n) \defeq \{ m \in \NN \,|\, m \geq n \}. \]

  \begin{block}{Proof}
    If~$M \subseteq \widehat\NN$ is an inhabited subset, its minimum is
    \[ \min M = \bigcup M \in \widehat\NN. \]
    The proof of the universal property is straightforward.
  \end{block}
}

\note{
  \begin{block}{External translation (see Mulvey's \emph{Intuitionistic algebra
  and representations of rings})}
    Let~$X$ be a topological space and consider the constant sheaf~$N$ with
    $\Gamma(U, N) = \{ f : U \to \NN \,|\ \text{$f$ continuous} \}$.
    Internally, the sheaf~$N$ plays the role of the ordinary natural numbers.
    Then there is an one-to-one correspondence:
    \begin{enumerate}
      \item Let~$A \hookrightarrow N$ be a subobject which is inhabited
      and upward closed from the internal point of view. Then
      \[
        x \longmapsto \inf\{ n \in \NN \,|\, n \in A_x \}
      \]
      is an upper semi-continuous function on~$X$.

      \item Let~$\alpha : X \to \NN$ be a upper semi-continuous function. Then
      \[ U \subseteq X \longmapsto \{ f : U \to \NN \,|\, \text{$f$
      continuous,\ \ $f \geq \alpha$ on~$U$} \} \]
      is a subobject of~$N$ which internally is inhabited and upward closed.
    \end{enumerate}
  \end{block}

  Under the correspondence, locally \emph{constant}
  functions map exactly to the [image of the] \emph{ordinary} internal natural numbers
  [in the completed natural numbers].
}

\note{
  \begin{itemize}
    \item Here is an explicit example of a completed natural number which is
    not an ordinary natural number: Let~$X =
    \Spec k[X]$ and~$\F = k[X]/(X-a)^\sim$. The rank of~$\F$ is~$1$ at~$a$ and
    zero elsewhere. It corresponds to the internal completed natural number
    \begin{multline*}
      z \defeq \min \{ n \in \NN \,|\, \speak{$\F$ can be generated by~$n$ elements} \} = \\
        \min\{ n \in \NN \,|\, n \geq 1 \vee \speak{the element $(X-a)$ of~$\O_X$
        is invertible} \}.
    \end{multline*}
    We have the internal implications
    \begin{align*}
      \Sh(X) \models \phantom{\neg}\,\speak{$(X-a)$ inv.} \Rightarrow z = 0\phantom{,} \\
      \Sh(X) \models \neg\,\speak{$(X-a)$ inv.} \Rightarrow z = 1,
    \end{align*}
    but we do \emph{not} have
    \begin{align*}
      \Sh(X) \models \speak{$(X-a)$ inv.} \ \vee\ \neg\,\speak{$(X-a)$ inv.},
    \end{align*}
    which would imply
    \begin{align*}
      \Sh(X) \models z = 0 \vee z = 1,
    \end{align*}
    i.\,e. the false statement that~$\F$ is locally free (of ranks~$0$ resp.~$1$).
  \end{itemize}
}

\note{
  \begin{itemize}
    \item \begin{justify}Here is a constructive proof of the statement that
    finitely generated vector spaces, for which the minimal number of
    generators is an actual natural numbers, are free:

    By assumption, the minimal number~$n \in \NN$ of generators for~$M$ exists.
    Let~$x_1,\ldots,x_n$ be a generating family of minimal length~$n$. We want
    to verify that it's linearly independent, so that it constitutes a basis.

    Let~$\sum_i \lambda_i x_i = 0$. If any~$\lambda_i$
    were invertible, the shortened family $x_1,\ldots,x_{i-1},$ $x_{i+1},\ldots,x_n$
    would also generate~$M$. By minimality of~$n$, this is not possible. So
    each~$\lambda_i$ is not invertible. By the field assumption on~$A$, it
    follows that each~$\lambda_i$ is zero.\end{justify}

    \item \begin{justify}In constructive mathematics, one can not show that
    every finitely generated vector space over a field admits a finite basis.
    (Exercise: Prove this by showing that this would imply the law of excluded
    middle.)
    This is not because the space might strangely turn out to be
    infinite-dimensional, but merely because one may not be able to explicitly
    exhibit a finite basis.\end{justify}
  \end{itemize}
}

@@

  \hil{Question:} How do the properties of
  \begin{itemize}
  \item an~$A$-module~$M$ in~$\Set$ and
  \item the~$\O_X$-\only<2->{module}\only<1>{\pointthis{module}{an important
  sheaf with~$M^\sim(X) = M$}}~$M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
  \end{itemize}
  \pause
  \vfill

  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
  \begin{itemize}
  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}.
  \end{itemize}
  Note: $M$ and $\underline{M}$ share all first-order properties.
  \pause
  \vfill

@@

\note{\scriptsize
  \begin{itemize}
    \item \begin{justify}Hakim worked out a theory of schemes internal to topoi (but
    without using the internal language) in her PhD thesis.\end{justify}

@@

    There is a similar story for local~$A$-algebras.
    See Wraith's \emph{Intuitionistic algebra: some recent developments in
    topos theory} for a short exposition on the usefulness of
    classifying topoi and universal models.\end{justify}
  \end{itemize}
}

>>>>

\title{First steps in synthetic algebraic geometry}
\author{Ingo Blechschmidt}
\date{February 7th, 2017}

@@

\newcommand{\pointthis}[2]{%
  \tikz[remember picture,baseline]{\node[anchor=base,inner sep=0,outer sep=0]%
    (#1) {#1};\node[overlay,rectangle callout,%
    callout relative pointer={(0.5cm,0.5cm)},fill=blue!20] at ($(#1.north)+(-1.1cm,-1.1cm)$) {#2};}%
}%

@@

  \hil{First steps in \\ synthetic algebraic geometry}
  \medskip

@@

  Algebraic Geometry Seminar \\
  University of Warwick \\
  \ \\
  Feburary 7th, 2017
  \par
\end{frame}

\frame[t]{\frametitle{Outline}\scriptsize\begin{itemize}\item[]\tableofcontents\end{itemize}}

@@

  \begin{changemargin}{2.5em}{2.5em}
    We describe how the internal language of certain toposes, the
    associated little and big Zariski toposes of a scheme, can be used
    to give simpler definitions and more conceptual proofs of the basic
    notions and observations in algebraic geometry.

    The starting point is that, from the "internal point of view" of the
    little Zariski topos, sheaves of rings and sheaves of modules look
    just like plain rings and plain modules. In this way, some concepts
    and statements of scheme theory can be reduced to concepts and
    statements of intuitionistic linear algebra. This simplifies working
    with sheaves and brings conceptual clarity.

    The internal language of the big Zariski topos goes even further. It
    incorporates Grothendieck's functor-of-points philosophy in order to
    cast modern algebraic geometry, relative to an arbitrary base
    scheme, in a naive language reminiscient of the classical Italian
    school. The base scheme looks like the one-element set from this
    point of view.

    The talk gives an introduction to this topos-theoretic point of view
    of algebraic geometry in the hope that it's useful to working
    algebraic geometers. No prior knowledge about toposes is supposed.
  \end{changemargin}
}

\section{The internal language of toposes}

\subsection{What is a topos?}
\frame[t]{\frametitle{What is a topos?}
  \begin{block}{Formal definition}A \hil{topos} is a category which has finite limits,
  is cartesian closed and has a subobject classifier.
  \end{block}

  \begin{block}{Motto}A topos is a category sufficiently rich to support
  an \hil{internal language}.
  \end{block}

  \begin{block}{Examples}\begin{itemize}
    \item \hil{$\Set$:} \tabto{1.4cm} category of sets
    \item \hil{$\Sh(X)$:} \tabto{1.4cm} category of set-valued sheaves on a space~$X$
    \item \hil{$\Zar(S)$:} \tabto{1.4cm} big Zariski topos of a base scheme~$S$
  \end{itemize}\end{block}

  \note{
    \begin{itemize}
      \item \begin{justify}While technically correct, the formal definition is actually
      misleading in a sense: A topos has lots of other vital structure, which
      is crucial for a rounded understanding, but is not listed in the
      definition (which is trimmed for minimality).

      A more comprehensive definition is: A \emph{topos} is a locally cartesian
      closed, finitely complete and cocomplete Heyting category which is exact,
      extensive and has a subobject classifier.\end{justify}

      \item See~\cite{leinster:topos} for a leisurely introduction to topos
      theory.
    \end{itemize}
  }
}

\subsection{What is the internal language?}
\frame[t]{\frametitle{What is the internal language?}
  The internal language of a topos~$\E$ allows to
  \begin{enumerate}
    \item construct objects and morphisms of the topos,
    \item formulate statements about them and
    \item prove such statements
  \end{enumerate}
  in a \hil{naive element-based} language:

  \begin{center}
    \small
    \begin{tabular}{ll}
      \toprule
      externally & internally to $\E$ \\
      \midrule
      object of~$\E$ & set \\
      morphism in~$\E$ & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      group object & group \\
      \bottomrule
    \end{tabular}
  \end{center}
}

@@

\note{
  \begin{itemize}
    \item \begin{justify}The rules are called \emph{Kripke--Joyal semantics} and can be
    formulated over any topos (not just sheaf toposes). They are not all
    arbitrary: Rather, they are very finely concerted to make the crucial
    properties about the internal language (see next slide) true.\end{justify}

@@

  \begin{block}{Crucial property: Soundness}
    If~$U \models \varphi$ and if $\varphi$
    implies $\psi$ \only<1>{constructively}\only<2->{\pointthis{constructively}{
      no $\varphi \vee \neg\varphi$,\ \
      no $\neg\neg\varphi \Rightarrow \varphi$,\ \
      no axiom of choice
    }},
    then~$U \models \psi$.
  \end{block}

  \vfill
  \vspace*{2em}

  \pause\pause
  \begin{block}{A first glance at the constructive nature}
  \begin{itemize}
    \item $U \models f = 0$
      \tabto{3.5cm} iff $f|_U = 0 \in \Gamma(U, \F)$.
    \item $U \models f = 0 \vee g = 0$
      \tabto{3.5cm} iff on a cover~$U = \bigcup_i U_i$,\\
      \tabto{3.5cm} \qquad\qquad $f|_{U_i} = 0$ or~$g|_{U_i} = 0$.
    \item $U \models \neg\neg(f = 0)$
      \tabto{3.5cm} iff~$f = 0$ on a dense open subset of~$U$.
  \end{itemize}
  \end{block}
}

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Why is constructive mathematics
  interesting?\end{center}
  \vspace{-1em}
  \begin{itemize}
    \item The internal logic of most toposes is constructive.

@@

\note{%
  \vspace{-0.5em}%
  \begin{center}\large\bf Three rumours about constructive mathematics\end{center}
  \vspace{-1em}

  \begin{enumerate}
    \item \begin{justify}There is a false rumour about constructive mathematics, namely that
    the term \emph{contradiction} is generally forbidden. This is not the
    case, one has to distinguish between
    \begin{itemize}
      \item a true proof by contradiction: ``Assume~$\varphi$ were false.
      Then \ldots, contradiction. So~$\varphi$ is in fact true.''
    \end{itemize}
    which constructively is only a proof of the weaker
    statement~$\neg\neg\varphi$, and
    \begin{itemize}
      \item a proof of a negated formula: ``Assume~$\psi$ were true. Then
      \ldots, contradiction. So~$\neg\psi$ holds.''
    \end{itemize}
    which is a perfectly fine proof of~$\neg\psi$ in constructive mathematics.\end{justify}

@@

\note{
  \begin{enumerate}
    \item[3.] \begin{justify}There is one last false rumour about constructive mathematics: Namely
    that most of mathematics breaks down in a constructive setting. This is
    only true if interpreted naively: Often, already very small changes to the
    definitions and statements (which are classically simply equivalent
    reformulations) suffice to make them constructively acceptable.

    In other cases, adding an additional hypotheses, which is classically
    always satisfied, is necessary (and interesting). Here is an example: In
    constructive mathematics, one can not show that any inhabited subset of the
    natural numbers possesses a minimal element. [One can also not show the
    negation -- recall the previous false rumour.] But one can show (quite
    easily, by induction) that any inhabited and \emph{detachable} subset
    of the natural numbers possesses a minimal element. A subset~$U \subseteq
    \NN$ is detachable iff for any number~$n \in \NN$, it holds that~$n
    \in U$ or~$n \not\in U$.

@@

  \begin{justify}See~\cite{stanford:constr,dalen:int,troelstra:dalen:constr}
  for references about constructive mathematics. The blog~\cite{bauer:blog} is
  also very informative.\end{justify}
}

\section{The little Zariski topos of a scheme}

@@

\begin{frame}\frametitle{Building a dictionary}
  \sloganwithoutborder{\hil{Understand notions of algebraic geometry
  as notions of algebra internal to~$\boldsymbol{\Sh(X)}$.}}
  \begin{center}
    \small
    \scalebox{0.93}{\begin{tabular}{ll}
      \toprule
      externally & internally to $\Sh(X)$ \\
      \midrule
      sheaf of sets & set \\
      morphism of sheaves & map of sets \\
      monomorphism & injective map \\
      epimorphism & surjective map \\
      \midrule
      sheaf of rings & ring \\
      sheaf of modules & module \\
      sheaf of finite type & finitely generated module \\
      finite locally free sheaf & finite free module \\
      % coherent sheaf & coherent module \\
      tensor product of sheaves & tensor product of modules \\
      sheaf of Kähler differentials & module of Kähler differentials \\
      sheaf of rational functions & total quotient ring of~$\O_X$ \\
      dimension of $X$ & Krull dimension of~$\O_X$ \\
      \bottomrule
    \end{tabular}}
  \end{center}

@@

@@

  \hil{Question:} How do the properties of
  \begin{itemize}
  \item an~$A$-module~$M$ in~$\Set$ and
  \item the~$\O_X$-module $M^\sim$ in~$\Sh(X)$, where~$X = \Spec A$, relate?
  \end{itemize}
  \pause
  \vfill

  \hil{Observation:} $M^\sim = \underline{M}[\F^{-1}]$, where
  \begin{itemize}
  \item $\underline{M}$ is the constant sheaf with stalks~$M$ on~$X$ and
  \item $\F \hookrightarrow \underline{A}$ is the \hil{generic prime filter}
  with stalk~$A \setminus \mathfrak{p}$ at $\mathfrak{p} \in \Spec A$.
  \end{itemize}
  Note: $M$ and $\underline{M}$ share all first-order properties.
  \pause
  \vfill

@@

\note{\scriptsize
  \begin{itemize}
    \item \begin{justify}Hakim worked out a theory of schemes internal to toposes (but
    without using the internal language) in her PhD thesis.\end{justify}

@@

    There is a similar story for local~$A$-algebras.
    See Wraith's \emph{Intuitionistic algebra: some recent developments in
    topos theory} for a short exposition on the usefulness of
    classifying toposes and universal models.\end{justify}
  \end{itemize}
}
